### 2025-09-25 paper/sec/conclusion.tex
conclusions and future work

\section{Conclusion}
\label{sec:conclusion}

\input{fig/nl-further-works}

>>>>

\section{Conclusions and Future Work}
\label{sec:conclusion}

We introduced a proof-of-concept system for authoring transparent, data-driven documents by combining LLM-based code synthesis with Fluid's provenance-tracking runtime. Our evaluation on SciGen shows that the approach can reliably link natural language claims to their underlying data, while also revealing common failure modes such as ambiguity and misleading input.

%\subsection{Future Work}
%\label{sec:conclusion:future-work}

Future work includes reducing reliance on predefined helper functions
such as \kw{growShrink} and \kw{trendWord}. While these provide a uniform
framework, we aim to enable the system to operate in their absence,
for instance by turning ``definition not found'' errors into augmented
prompts that trigger automatic generation of missing definitions. We
also plan to broaden the scope of supported artifacts, extending
interpretation to visualisations and intermediate datasets derived
from cleansing or aggregation, and to cover additional idioms such as
cardinals, multiplicatives, rounding, and graded adjectives.

Another priority is improving integration and validation. Embedding
the system into developer and authoring environments such as VSCode or
Cursor would make the workflow more seamless, while automatic
generation of counterfactual test cases could strengthen validation at
authoring time. Finally, distinguishing between \emph{referential terms} with
fixed denotations and queries with data-dependent values may help in
repairing false or inconsistent statements, ensuring that generated
expressions remain aligned with both the data and the author's intent.

%Providing helper functions such as \kw{growShrink} and \kw{trendWord} as a predefined library is in some ways
%beneficial because they offer a uniform framework for interpreting a given scholarly document. We would also
%like make the system able to operate in the absence of these, perhaps adding a loopback condition to turn
%``Definition not found'' errors into an augmented prompt to generate the appropriate definition.

%Another important direction is to broaden the scope of the artifacts than can be referred to and the natural
%language forms that can be used to refer to them. For example, many claims in scholarly writing refer to
%visualisations and intermediate data sets derived via cleansing, imputation or aggregation. These should also
%be made available to the LLM (for visualisations, as structured data rather than as static bitmaps), enabling
%interpretation of natural language in terms of visual information and other intermediate analyses. Extending
%coverage to additional idioms beyond those considered in Table~\ref{tab:natural-language-forms}, such as
%cardinals and multiplicatives (\textfrag{three} and \textfrag{three times}), rounding (\textfrag{about 2
%degrees warmer}) and graded adjectives \textfrag{extremely likely}, would also expand the expressivity of the
%assistant.

%Consider the false statement \textfrag{The population of France is the largest in the EU}. There are at least
%two ways to repair this statement so that it comes out true: treat the subject (France) as fixed, but adjust
%the predicate so that it matches the facts (\textfrag{second largest}), or treat the predicate as fixed, but
%keep adjust the subject to match (\textfrag{Germany}). To make these sorts of (pragmatically important)
%distinctions, we may need to introduce the idea of \emph{referential terms} that have fixed denotations, in
%contrast to queries, whose values are contingent on data.

%Finally, integration into developer and authoring environments such as VSCode or Cursor would make the
%workflow more seamless for authors. Automatic generation of the counterfactual test cases discussed in
%\secref{validation-with-counterfactual-data} could further strengthen validation at authoring-time, improving
%the trustability of interpreted claims.

### 2025-09-25 paper/sec/evaluation.tex
conclusions and future work

\begin{figure}
    \centering
    \begin{subfigure}{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fig/success_rate_by_category}
        \caption{Success rate by Linguistic Category}
        \label{fig:success_rate_by_category}
    \end{subfigure}\hfill
    \begin{subfigure}{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fig/success_rate_by_complexity}
        \caption{Success rate by Complexity}
        \label{fig:success_rate_by_complexity}
    \end{subfigure}
    \caption{Success rate of the proposed system, measured over 5 runs with \gptfour}
    \label{fig:success_rate_comparison}
\end{figure}

@@

Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and without target-value sharing.
The results show that providing the target value is crucial for achieving a high success rate.
Moreover, the \InterpretationAgent generates a correct expression in 74.9\% of the cases (S.D. 3.0\%).

The system achieved performance above 68\% in comparison, 77.3\% in data retrieval, and 97\% in min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (0\%).

Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 62\% for simple expressions (one category), increased to 91\% for two categories,
while no correct expression was generated for expression with 3 categories.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 57.1\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance

>>>>

\begin{figure}
    \centering
    \begin{subfigure}{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fig/success_rate_by_category}
        \caption{Success rate by Linguistic Category}
        \label{fig:success_rate_by_category}
    \end{subfigure} \hfill
    \begin{subfigure}{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fig/success_rate_by_complexity}
        \caption{Success rate by Complexity}
        \label{fig:success_rate_by_complexity}
    \end{subfigure}
    \caption{Success rate of the proposed system, measured over 5 runs with \gptfour}
    \label{fig:success_rate_comparison}
\end{figure}

@@

Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and without target-value sharing.
The results show that providing the target value is crucial for achieving a high success rate.
Moreover, the \InterpretationAgent generates a correct expression in approximately XX\% of the cases.

The system achieved performance above 70\% in comparison, data retrieval, and min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (50\%).

Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 75\% for simple expressions (one category), increased to 85\% for two categories,
and dropped to 58\% when three categories were involved.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 39.4\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance
